Formal Modelling of a Microcontroller Instruction Set in B

### 2009-06-26 trunk/papers/SBMF2009/source/paper.tex
Initial import.

This work describes briefly the modelling of microcontroller Z80~\cite{Z80_manual} and some techniques used to build
it. This modelling\footnote{The interested reader in more details is invited to visit our repository
at: http://code.google.com/p/b2asm} use the B method~\cite{Abrial}, because it has a great maturity and advanced
support. Besides, the modelling of microcontroller can extend the level of verification of B method, as it will be
explained.

The B method supports the construction of safety systems model by verification of proofs that guarantees its
correctness. So, an initial abstract model of the system is defined and refined until the implementation model. The
B tools  still have the code generator to programming language, but this feature is not verified. In~\cite{Dantas_SBMF08} is presented an approach to extend the verification of B method until the assembler language.
In this approach, the construction of systems begins from the abstract level and it is formally refined until the
target assembly platform.

Therefore, the first step of this approach is to build the platform modelling, of the microcontroller, in B. Using the
responsibility division mechanism provided by B, basic modules was developed to be used as elementary library to
help the constructing of microcontroller model. The elementary library has many definitions about common concepts used
in the microcontrollers. This library is used by Z80 model and other two model microcontrollers that
are under construction.

The microcontroller modelling can be used in platform projects documentation, simulators construction, consistency
verification and software verification at assembly level. The verification at assembly level is more interesting
utility because it allows the development of trustworthy software by construction, that is extremely desirable in
embedded critical software.

In this paper we introduced the modelling of Z80 microcontroller, since elementary library until assembly
instructions. The paper is structured as follows. Section~\ref{sec:B_method} provides a general introduction to the
B method. Section~\ref{sec:models} presents the elementary libraries and of microcontrollers basic concepts.
Section~\ref{sec:z80} presents the Z80 B modelling. Section ~\ref{sec:Proofs} explain a little about the proofs of
models. The next section (\ref{sec:relatedworks}) discusses the related works. Finally, the last section has the
conclusions.

\section{B Method}
\label{sec:B_method}

@@

 Essentially, a B module contains two main parts: a header and the available operations, in the
 figure~\ref{fig:maqB} has a intuitive example. The clause $\mathit{MACHINE}$ has the name of module.  The following two are used to access an external module and create an instance of an external module. The
 $\mathit{VARIABLES}$  declares the variable names. The following declares its types and the restrictions. The
 $\mathit{INITIALIZATION} $  defines the its initial value. The last define the operations that remember procedures
 from programming language.

\begin{figure}
%\begin{small}
$$
\begin{array}{l}
\MACHINE \quad  \mathit{micro} \quad \quad \quad \quad \quad  \quad \quad \quad \quad \INITIALISATION  \mathit{pc}
:= 0\\
\SEES \quad \mathit{TYPES}, \mathit{ALU} \quad \quad \quad \quad \quad \quad \quad \quad \OPERATIONS\\
\INCLUDES    \quad \mathit{MEMORY} \quad \quad \quad \quad \quad \quad \quad \mathit{JMP} \mathit{( jump )} =
\\
\VARIABLES \quad    \mathit{pc}  \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \PRE \mathit{jump}
\in\mathit{INSTRUCTION}\\ 
 \INVARIANT \quad \mathit{pc} \in \mathit{INSTRUCTION}  \quad \ \THEN \mathit{pc} :=
\mathit{jump}\\  
\quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad
\END \quad  \ldots \\
% \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad
%\quad  \ldots
\end{array}
$$
% \begin{array}{l}
% \MACHINE \quad \quad \quad \quad \quad \quad  \quad \quad \quad \quad \INITIALISATION \\
% \quad   \mathit{micro} \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad   \quad   \mathit{pc} := 0\\
% \SEES \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad   \quad   \OPERATIONS\\
% \quad \mathit{TYPES}, \mathit{ALU}\quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad   \quad \\
% \INCLUDES                \quad \quad \quad \quad \quad \quad \quad \quad \quad   \quad   \mathit{JMP} \mathit{( jump )} = \\
% \quad   \mathit{MEMORY}   \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \PRE\\
% \VARIABLES      \quad \quad \quad \quad \quad \quad \quad \quad \quad  \quad \quad  \mathit{jump} \in
% \mathit{INSTRUCTION}\\ \quad   \mathit{pc}    \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad\quad \quad \THEN\\
% \INVARIANT   \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad   \mathit{pc} := \mathit{jump}\\
% \quad \mathit{pc} \in \mathit{INSTRUCTION} \ \ \quad \quad \quad \quad \quad \quad \END\\
% \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad
% \quad \quad  \ldots
% \end{array}

%\end{small}
\caption{A part of B simplified module}
\label{fig:maqB}
\end{figure}

\section{Overview of Plataform's B Models}
\label{sec:models}

We have been developed a reusable set of basic definitions to model hardware concepts and data types
concepts. These definitions are grouped in two projects of libraries and other project to model the
platform. Thus, our workspace is arranged through three projects: hardware library, types library and
the project of specific platform, in this case the Z80.

The dependency graph between these projects is depicted in figure~\ref{fig:hardware-definition-graph}
and the next sections will explain each one of them.

\begin{figure}[h] \centering
\includegraphics[width=.92\textwidth]{diagramaEstrutural_vertical.jpg}
 \caption{Dependency graph between the projects that compose the platform model of Z80.}
\label{fig:hardware-definition-graph}
\end{figure}

\subsection{Definitions to Represent and Manipulate Bits}

@@

\subsection{Modelling Bytes and Bit Vector of Length 16}

@@

The $\mathit{BYTE\_INDEX}$ is domain of byte modelled. It starts at 1 to obey a definition of sequences from B
method. However, the functions from byte encapsulate the access and these functions use the
$\mathit{PHYS\_BYTE\_INDEX}$. The $\mathit{BYTE}$ type is a specialized type from $\mathit{BIT\_VECTOR}$, but it has
a size limit. Some others more specific definitions are created to help the developer. The type $\mathit{BV16}$ is
created for bit vector of length 16 in a similar way.

@@

Bit vectors are used to represent and combine numbers: integer ranges (signed or unsigned). Then, our library
defines functions to manipulate this date, for example, the function $\mathit{bv\_to\_nat}$ that maps bit
vectors to natural numbers:

@@

 Usually, for each type module just needs to detail or redefines the pre-defined concepts. For
 example, the function $\mathit{bv\_to\_nat}$ from bit vector arithmetics that is specialized to
 $\mathit{byte\_uchar}$. The set $\mathit{BYTE}$ is a subset of the
$\mathit{BIT\_VECTOR}$, then this function can defined as follows:

@@

The definitions of the library types reuse the basic definitions from the hardware library. This provides greater
confidence and facilitates the prove process, because the prover can reuse the
previous defined lemma.

The inverse function is easily defined as $\mathit{uchar\_byte}$.

@@

The \textit{Z80} is an important CISC microcontroller developed by \textit{Zilog}~\cite{Z80_manual}. It supports
158 different instructions. These instructions are classified into these categories: load and exchange; block transfer
and search; arithmetic and logical; rotate and shift; bit manipulation (set, reset, test); jump, call and return; input/output; and basic cpu control.

The main module includes an instance of memory module and accesses the definitions from: basic data types modules
and \textit{ALU}.

@@

% Chacteristics of microcontroller
Each instruction is represented by B operations in the module Z80. By default, all parameters from
operations are or predefined elements in the model or integers values in the decimal representation.
The internal registers contain 208 bits of reading/writing memory. It includes two sets of six general purpose
registers which may be used individually as 8-bits registers or as 16-bits register pairs.  The work registers are
represented by variable $\mathit{rgs8}$. The domain of $\mathit{rgs8}$ ($\mathit{id\_regs8}$) is a set formed by
identifiers of registers of 8 bits. These registers can be accessed in pairs, forming 16-bits, resulting in other
set of identifiers of 16-bits registers, named $\mathit{id\_reg16}$. The main work register of Z80 is the
accumulator ($\mathit{rgs8(a0)}$) used for arithmetic, logic, input/output and loading/storing operations.
\subsection{Modelling registers, input and output ports and instructions}

The Z80 has many register of diferrents types and many instructions. It includes alternative set of accumulator,
flag and general registers. The CPU contains a stack pointer ($\mathit{sp}$), program counter ($\mathit{pc}$), two
index registers ($\mathit{ix}$ and $\mathit{iy}$), an interrupt register ($\mathit{i\_}$), a refresh register
($\mathit{r\_}$), two bits ($\mathit{iff1}$, $\mathit{iff2}$) used to control the interruptions, a pair of bits to
define the interruption mode ($\mathit{im}$) and the input and output ports ($\mathit{i\_o\_ports}$). Below, its
definitions are specified in the \textit{INVARIANT}.

\begin{sloppypar}
\bf INVARIANT

\hspace*{0.10in}\it rgs8  $\in$  \it id\_reg\_8  $\fun$  \it BYTE  $\land$ pc  $\in$  \it INSTRUCTION  $\land$

@@

% [Conclusao] This project has many proof obligations and
The building of model was a hard and difficult task, because we needed to build, prove, fix and re-prove
several predicates of the libraries.

However, few proof commands\footnote{The proof commands are step that just indicate a walk to prover make the proof,
thus theses commands cannot introduce false hypotheses.} can be useful to prove the majority of the proof
obligations. Because, there are many similar assembler instructions. A good example is a set of 17 proof commands
that quickly aided the verification of 99\% (2295) of proofs WD. Finally, the techniques mentioned  help
the verification of all the 2926 proof obligations.

@@

% [ I need review the english]

There are in the literature of computer science some approaches~\cite{BHDL_2003,GEMPLUS_99} to model hardware and
the virtual machines using the B method. Then, the B method has been used successfully to model the operational
semantic. However the cost of modelling was still expensive and this paper quoted some techniques to lower the cost
of modelling. Although, the modelling of the Z80 do not have modelled some small aspects.
% The objective of researchers

In general, the researchers~\cite{GEMPLUS_99} are concentrated at model the Java Virtual Machines (JVM), but it is
not easy to specify, because it is very big and complex. Then, this work specify a model more simple to get better
results, at the same time, the model stay near to model of actual microcontrollers. The main motivation of our
research is the verification of assembler level and consequently the verification of hardware model. Thus, the some
aspects are not modelled. For example, the time of execution of instruction, the pipeline and others questions not
directly related with the semantic software. However, there are many other specialized techniques to verify these
questions.

 \section{Conclusions}
\label{sec:conclusions} 
This work has shown an approach to the formal modelling of microcontrollers using the B method. This approach
provide interesting results and had a great automatization in the prove process. For instance, it is interesting to
say that: during the building process of the Z80 model we found some errors and ambiguities in the official
manual\cite{Z80_manual}. The platform model can replace (or improve) the documentation used by assembler
programmers. Because the formal model represents the instructions effects and the B method has an easy notation,
similar to pascal language.  Besides, the formal model restricts the definitions to correct typing, uses expressions
well-defined and allows verifying properties of the model.

@@

 \paragraph{Acknowledges:}
This work received supports from ANP (National Agency of Oil) and CNPq.
% Besides, the researcher Bruno Gomes help in the revision.

>>>>

The B method~\cite{Abrial} supports the construction of safety systems
model by verification of proofs that guarantees its correctness. So,
an initial abstract model of the system requirements is defined and
then it is refined until the implementation model. Development
environments based on the B method also include source code generators
for several programming languages, but the result of this translation
cannot be compared by formal means. \cite{Dantas_SBMF08} presented
recently an approach to extend the scope of the B method up to the
assembly level language language. One key component of this approach
is to build, within the framework of the B method, formal modelos of
the instruction set of such assembly languages.

This work gives an overview of the formal modelling of the instruction
set of the Z80 microcontroller~\cite{Z80_manual}\footnote{The
  interested reader in more details is invited to visit our repository
  at: http://code.google.com/p/b2asm.}. Using the responsibility
division mechanism provided by B, auxiliary libraries of basic modules
were developed as part of the construction of microcontroller
model. Such library has many definitions about common concepts used in
the microcontrollers; besides the Z80 model, it is used by two other
microcontrollers models that are under way.

Other possible uses of a formal model of a micro-controller
instruction set include documentation, the construction of simulators,
and be possibly the starting point of a verification effort for the
actual implementation of a Z80 design. Moreover the model of the
instruction set could be instrumented with non-functional aspects,
such as the number of cycles it takes to execute an instruction, to
prove lower and upper bounds on the execution time of a routine. The
goal of this project, though, is to provide a basis for the generation
of software artifacts at the assembly level that are amenable to 
refinement verification within the B method.

This paper is focused on the presentation of the Z80 model, including
elementary libraries to describe hardware aspects. The paper is
structured as follows. Section~\ref{sec:B_method} provides a short
introduction to the B method. Section~\ref{sec:models} presents the
elementary libraries and the modelling of some elements common to
microcontrollers.  Section~\ref{sec:z80} presents the B model of the
Z80 instruction set. Section ~\ref{sec:Proofs} provides some
information on the proof effort needed to analyze the presented
models. Related work is discussed in
Section~\ref{sec:relatedworks}. Finally, the last section is devoted
to the conclusions.

\section{Introduction to the B Method}
\label{sec:B_method}

@@

Essentially, a B module contains two main parts: a header and the
available operations. Figure~\ref{fig:maqB} has a very basic
example. The clause $\mathit{MACHINE}$ has the name of module.  The
following two clauses respectively reference external modules and
create an instance of an external module. The $\mathit{VARIABLES}$
declares the name of the variables that compose the state of the
machine. Next, the $\mathit{INVARIANT}$ clause defines the type and
other restrictions on the variables. The $\mathit{INITIALIZATION} $
sepecifies the initial states. Finally, operations correspond to
the transitions between states of the machine.

\begin{figure}
  % \begin{small}
  $$
  \begin{array}{lcl}
    \begin{array}[t]{l}
      \MACHINE \quad  \mathit{micro} \\
      \SEES \quad \mathit{TYPES}, \mathit{ALU} \\
      \INCLUDES  \quad \mathit{MEMORY} \\
      \VARIABLES \quad    \mathit{pc}  \\
      \INVARIANT \quad \mathit{pc} \in \mathit{INSTRUCTION} 
    \end{array}
    & \hspace*{1cm} &
    \begin{array}[t]{l}
      \INITIALISATION  \mathit{pc} := 0 \\
      \OPERATIONS\\
      \mathit{JMP} \mathit{( jump )} = \\
      \quad \PRE \mathit{jump} \in \mathit{INSTRUCTION}\\ 
      \quad \THEN \mathit{pc} := \mathit{jump}\\  
      \quad \END \\
      \END
    \end{array}
  \end{array}
  $$
\caption{A very basic B machine.}
\label{fig:maqB}
\end{figure}

\section{Model structure and basic components}
\label{sec:models}

We have been developed a reusable set of basic definitions to model
hardware concepts and data types concepts. These definitions are
grouped into two separate development projects and are available as
libraries. A third project is devoted to the higher-level aspects of
the Z80 model. Thus, the workspace is composed of: a hardware library,
a types library and a project for the specific platform, in this case
the Z80. The corresponding dependency diagram is depicted in
Figure~\ref{fig:hardware-definition-graph}; information specific to
each project is presented in the following.

\begin{figure}[h] \centering
\includegraphics[width=.92\textwidth]{diagramaEstrutural_vertical.jpg}
 \caption{Dependency diagram of the Z80 model.}
\label{fig:hardware-definition-graph}
\end{figure}

\subsection{Bit representation and manipulation}

@@

\subsection{Modelling Bytes and Bit Vectors of Length 16}

@@

The $\mathit{BYTE\_INDEX}$ is the domain of the functions modelling
bytes. It starts at 1 to obey a definition of sequences from B
method. However, it is common in hardware architectures to start
indexing from zero. The definition $\mathit{PHYS\_BYTE\_INDEX}$ is
used to provide functionalities obeying this convention. The
$\mathit{BYTE}$ type is a specialized type from
$\mathit{BIT\_VECTOR}$, but it has a size limit. Other specific
definitions are provided to facilitate further modelling: the type
$\mathit{BV16}$ is created for bit vector of length 16 in a similar
way.

@@

Bit vectors are used to represent and combine numbers: integer ranges
(signed or unsigned). Therefore, our library includes functions to
manipulate such data, for example, the function $\mathit{bv\_to\_nat}$
that maps bit vectors to natural numbers:

@@

Usually, each type module just needs to instantiate concepts that were
already defined in the hardware modelling library.  For example, the
function $\mathit{bv\_to\_nat}$ from bit vector arithmetics is
specialized to $\mathit{byte\_uchar}$. As the set $\mathit{BYTE}$ is a
subset of the $\mathit{BIT\_VECTOR}$, this function can defined
as follows:

@@

The definitions of the library types reuse the basic definitions from
the hardware library. This provides greater confidence and facilitates
the proof process, because the prover can reuse the previously defined
lemma.

The inverse function $\mathit{uchar\_byte}$ is easily defined:

@@

The \textit{Z80} is a CISC microcontroller developed by
\textit{Zilog}~\cite{Z80_manual}. It supports 158 different
instructions. These instructions are classified into these categories:
load and exchange; block transfer and search; arithmetic and logical;
rotate and shift; bit manipulation (set, reset, test); jump, call and
return; input/output; and basic cpu control.

The main module includes an instance of the memory module and accesses the
definitions from basic data types modules and the \textit{ALU} module.

@@

% Chacteristics of microcontroller
Each instruction is represented by a B operation in the module Z80. By
default, all parameters from operations are either predefined elements
in the model or integers values in the decimal representation.  The
internal registers contain 208 bits of reading/writing memory. It
includes two sets of six general purpose registers which may be used
individually as 8-bits registers or as 16-bits register pairs.  The
working registers are represented by variable $\mathit{rgs8}$. The
domain of $\mathit{rgs8}$ ($\mathit{id\_regs8}$) is a set formed by
identifiers of registers of 8 bits. These registers can be accessed in
pairs, forming 16-bits, resulting in another set of identifiers of
16-bits registers, named $\mathit{id\_reg16}$. The main working
register of Z80 is the accumulator ($\mathit{rgs8(a0)}$) used for
arithmetic, logic, input/output and loading/storing operations.

\subsection{Modelling registers, input and output ports and instructions}

The Z80 has different types of registers and instructions. The CPU
contains general-purpose registers ($\mathit{id\_reg\_8}$), a stack
pointer ($\mathit{sp}$), program counter ($\mathit{pc}$), two index
registers ($\mathit{ix}$ and $\mathit{iy}$), an interrupt register
($\mathit{i\_}$), a refresh register ($\mathit{r\_}$), two bits
($\mathit{iff1}$, $\mathit{iff2}$) used to control the interruptions,
a pair of bits to define the interruption mode ($\mathit{im}$) and the
input and output ports ($\mathit{i\_o\_ports}$). Below, part of the
corresponding definitions are replicated from the \textit{INVARIANT}:

\begin{sloppypar}
\bf INVARIANT
\hspace*{0.10in}\it rgs8  $\in$  \it id\_reg\_8  $\fun$  \it BYTE  $\land$ pc  $\in$  \it INSTRUCTION  $\land$

@@

% [Conclusao] This project has many proof obligations and
The construction of then model was a difficult task, as several
iterations were needed to provide the good library definitions 
as well as to optimize the definition of the functionality of
the microprocessor instructions by grouping common data
manipulation into auxiliary functions.

However, few proof commands\footnote{The proof commands are steps that
  direct the prover to find the proof, and cannot introduce false
  hypothesis.} need to be used to prove most proof obligations. As
there are many similar assembly instructions, some human-directed
proofs, when replayed, could discharge other proof obligations. A good
example is a set of 17 proof commands that quickly aided the
verification of 99\% (2295) of proofs WD. Finally, all of the 2926
proof obligations were proved using the tool support of the
development environment.

@@

There are in the literature of computer science some
approaches~\cite{BHDL_2003,GEMPLUS_99} to model hardware and the
virtual machines using the B method. Then, the B method has been used
successfully to model the operational semantic. However the cost of
modelling was still expensive and this paper quoted some techniques to
lower the cost of modelling. 
% Although, the modelling of the Z80 do not
%have modelled some small aspects.

% The objective of researchers

In general, the researchers employing the B method have focused on
more abstract level of description of software.  Considering low-level
aspect, there has been previous work on modelling the Java Virtual
Machine\cite{GEMPLUS_99}.

The main motivation of our research is the development of verified
software up to the assembly level, which requires specifying the
semantics of the underlying hardware. Thus, some aspects were not
modelled in our work such as the execution time of the instructions.
Also we did not the micro-architecture of the hardware as the scope of
our work does not include hardware verification. However, there are
many other specialized techniques to verify these questions.

 \section{Conclusions}
\label{sec:conclusions}

This work has shown an approach to the formal modelling of the
instruction set of microcontrollers using the B method. During the
construction of this model, several ambiguities and errors were
encountered in the during the official reference~\cite{Z80_manual}. As
the B notation has a syntax that is not too distant from that of
imperative programming languages, such model could be used to improve
the documentation used by assembler programmers. Besides, the formal
notation used is analyzed by software that guarantees the
correctness of typing, the well-definedness of expressions, in addition
to safety properties of the micro-controller state.

@@

\paragraph{Acknowledges:}
This work received support from ANP (\textit{Ag\^{e}ncia Nacional do
  Petr\'{o}leo, G\'{a}s Natural e Biocombust\'{\i}veis}) and CNPq
(\textit{Conselho Nacional de Desenvolvimento Cient\'{\i}fico e
  Tecnol\'{o}gico}).
